Over Twenty Years of Virtual Machine Development and Debugging Through Simulation

### 2018-08-15 paper.tex
added 4.1

TO WRITE FROM MY YOUTUBE VIDEO.
Back-in-time debugging of machine state
Conditional stepping.

Ref Sista ManLang
\cite{SistaFastStartUp}
END TODO

>>>>

One idea we implemented in the recent year was to support a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartUp}. To be able to generate efficient bytecodes, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}. For each new bytecode, we introduced new templates in the JIT to generate efficient machine code for the optimized methods. After all the basic unit tests worked, we ran the speculative optimizer and got a crash. We could figure out which bytecode method was triggering the crash, but we had no idea from which template it came from, in addition, optimized methods include many inlined methods and are very large, so it's difficult to figure out where the issue comes from looking at thousands of bytecodes.

We created a snapshot where the faulty method was executed right after start-up. We started the VM simulator, and set-up a conditional breakpoint so that simulator would stop when the JIT would generate that method to machine code. Then, when the simulator effectively stopped, we changed the conditional breakpoint to stop execution where the address corresponding to the faulty method entry in machine code would be used, either by a call from the interpreter or through inline cache relinking. The simulator stopped again, about to execute the machine code corresponding to the faulty method. We then cloned the simulator to be able to reproduce the crash again and again.

Executing the faulty method led to an assertion failure. However, that assertion failed in a store check, telling us that the object to store into looked suspicious (outside of the heap). It happens that this object was read from a field on stack, and that this field held an incorrect address. We could not tell anymore at this point in the execution what instruction wrote on stack the invalid address. So we discarded the cloned simulation, and cloned a fresh simulator again, just before the execution of the faulty method in machine code. This time, we enabled single stepping (\emph{i.e.,} the processor simulator simulates one instruction at a time) and we added a breakpoint stopping execution when the specific field on stack would be written to an incorrect value. In this case, the breakpoint was checked in between each machine instruction, and execution stopped right after the machine instruction which wrote the incorrect value on stack. From the machine instruction address, we could map it back to the bytecode template (it mapped to the new bytecode template for inlined allocations), built a simpler method crashing the runtime and fixed the template using in-image compilation (See next subsection).

The debugging process used here is exampled in a youtube video \footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.
